Federated learning enables collaborative training without centralising raw data,
but it remains vulnerable to Byzantine clients and model poisoning. Robust
aggregation is especially difficult under non-IID data, where benign clients may
have persistent and legitimate update differences, and under temporal attacks
that create slow cross-round drift.

This dissertation proposes \emph{HRAC} (History-Residual Adaptive Clipping), an
oracle-free aggregation method that performs bounded suppression rather than
hard exclusion. HRAC combines an EMA-smoothed median norm cap with per-client
history-residual clipping, then applies a delayed soft weighting rule based on
residual-change statistics. By comparing clients primarily with their own update
histories, HRAC aims to reduce harmful influence without forcing benign
non-IID clients into a single global cluster.

Experiments on CIFAR-10 with 10 clients and $30\%$ Byzantine participation show
that HRAC remains competitive under benign training and improves robustness in
several structured or temporally persistent attack settings, particularly
non-IID MSA, HisMSA, ALIE, and IPM. Component ablations indicate that the global
median cap is especially influential under MSA, while $\nu$-based temporal
weighting contributes most clearly under ALIE. The results support the use of lightweight client-specific
residual histories, while also showing limitations under protocol-compliant
label poisoning, bit flipping, and limited experimental coverage.
