\doublespacing % Do not change - required

\chapter{Convergence Analysis of HRAC}
\label{app:convergence-analysis}

\thesisspacing % Do not change - required

This appendix gives the conditional convergence argument supporting the brief
statement in Chapter~\ref{ch:method-implementation}. The proof treats HRAC as a
robust aggregation rule with bounded clipping distortion, bounded history-state
growth, and bounded residual Byzantine weight mass, then inserts the resulting
aggregation error into a standard smooth non-convex descent argument. It should
be read as a bounded-influence stability analysis under stated assumptions, not
as a proof that HRAC identifies Byzantine clients in all settings.
\section{Introduction}

The goal is to separate the proof into two parts. Rather than proving the code
line by line, HRAC is abstracted as a robust aggregation rule and analysed as
follows:

\begin{enumerate}[label=\arabic*.]
    \item show that the HRAC aggregate remains close to the ideal honest
    conditional centre;
    \item show that such an aggregation error can be absorbed into the descent
    analysis of smooth non-convex optimisation.
\end{enumerate}

For the adaptive statistics, the main theorem-level target is not universal
pointwise ordering between Byzantine and honest clients at every late-stage
round. Instead, the primary claim is the weaker and empirically better-matched
one: Byzantine clients exhibit a larger late-stage group-mean \(\nu\)-band,
which induces stronger penalty pressure and smaller weights once the activation
threshold becomes active. In particular, the intended mechanism is not that
honest statistics vanish while Byzantine statistics remain nonzero. Rather,
both groups may stay in positive late-stage bands, while Byzantine clients are
typically elevated relative to the honest population. Pointwise separation is
kept as a stronger sufficient refinement.

The clipping argument is stated at the level needed for this dissertation. To
make the global cap compatible with the clipped-moment lemma of Sadiev
et al.~\cite{sadiev2023highprob}, HRAC uses a lagged history-EMA median scale
as its global clipping radius. Thus the global threshold is
\(\F_t\)-measurable at round \(t\), just like the residual threshold determined
by the HRAC history state. The price of this proof route is a strong bounded
honest-update assumption, so the result certifies bounded-influence behaviour
rather than a full distribution-free FL guarantee.

\section{Problem Setup}

Let \(x_t \in \R^d\) denote the global model at round \(t\).
Suppose there are \(N\) participating clients, partitioned into an honest set
\(\HH\) and a Byzantine set \(\BB\), with
\[
    |\HH| = H, \qquad |\BB| = B, \qquad N = H + B.
\]

At round \(t\), the server receives client messages
\[
    \{\Delta_i^t\}_{i=1}^N \subset \R^d,
\]
where \(\Delta_i^t\) denotes a gradient-like descent direction, as in
Chapters~\ref{ch2} and~\ref{ch:method-implementation},
and performs the global update
\begin{equation}
    x_{t+1} = x_t - \eta_t g_t,
    \label{eq:server_update}
\end{equation}
where \(g_t\) is the aggregate produced by HRAC.

The optimization proof below uses the conditional honest centre as the
reference object, because the clipped-moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob} is stated conditionally on
\(\F_t\). For honest clients, define
\begin{equation}
    m_i^t:=\E[\Delta_i^t\mid\F_t],
    \qquad i\in\HH,
    \label{eq:honest_conditional_centre}
\end{equation}
and the average honest conditional centre
\begin{equation}
    \bar m_{\HH}^t
    :=
    \frac1H\sum_{i\in\HH}m_i^t.
    \label{eq:honest_centre_average}
\end{equation}

The objective is to minimise a possibly non-convex function
\[
    f : \R^d \to \R.
\]

\section{HRAC Aggregation Rule}

HRAC operates in three stages.

\subsection{Global norm clipping}

Define the current round median norm
\begin{equation}
    M_t := \med_{1\le j\le N}\norm{\Delta_j^t}.
    \label{eq:round_median_norm}
\end{equation}
HRAC maintains a history-EMA global scale
\begin{equation}
    s_{t+1}
    =
    \rho_gs_t+(1-\rho_g)M_t,
    \qquad 0<\rho_g<1,
    \label{eq:global_scale_ema}
\end{equation}
and uses the predictable global clipping cap with a fixed positive floor
\begin{equation}
    B_t := c_g\max\{s_t,s_{\min}\},
    \qquad c_g\ge1,\quad s_{\min}>0.
    \label{eq:global_cap}
\end{equation}
Thus \(B_t\in\F_t\): the current round median \(M_t\) updates the next round's
radius but is not used to clip the current round's updates. The globally
clipped update is
\begin{equation}
    \bar{\Delta}_i^t
    :=
    \clip(\Delta_i^t, B_t)
    :=
    \begin{cases}
        \Delta_i^t, & \norm{\Delta_i^t} \le B_t,\\[2mm]
        B_t \dfrac{\Delta_i^t}{\norm{\Delta_i^t}}, & \norm{\Delta_i^t} > B_t.
    \end{cases}
    \label{eq:global_clip}
\end{equation}
The implementation warm-starts the EMA scale from the first observed median and
then applies \eqref{eq:global_scale_ema}. In the analysis below, the counted
rounds start after this warm-start state is available, so the global cap used
inside the proof is predictable. Equivalently, one may view the first
current-median cap as a finite calibration prefix, which has no effect on the
asymptotic neighbourhood.

\subsection{Per-client history-residual clipping}

For each client \(i\), HRAC maintains a history state
\[
    (b_i^t,\mu_i^t,\nu_i^t),
\]
where \(b_i^t\) is a running centre, \(\mu_i^t\) is a scale estimate, and
\(\nu_i^t\) is a change statistic.

Define the residual
\begin{equation}
    r_i^t := \bar{\Delta}_i^t - b_i^t,
    \label{eq:residual}
\end{equation}
and the adaptive residual clipping threshold
\begin{equation}
    \tau_i^t := c \mu_i^t.
    \label{eq:tau}
\end{equation}

The clipped residual is
\begin{equation}
    \hat{r}_i^t
    :=
    \clip(r_i^t, \tau_i^t)
    =
    \begin{cases}
        r_i^t, & \norm{r_i^t} \le \tau_i^t,\\[2mm]
        \tau_i^t \dfrac{r_i^t}{\norm{r_i^t}}, & \norm{r_i^t} > \tau_i^t.
    \end{cases}
    \label{eq:residual_clip}
\end{equation}

The post-processed message becomes
\begin{equation}
    \tilde{\Delta}_i^t := b_i^t + \hat{r}_i^t.
    \label{eq:delta_tilde}
\end{equation}

The history centre \(b_i^t\) follows the implemented EMA dynamics:
for a newly appearing client, \(b_i^t\) is initialized by
\(\bar{\Delta}_i^t\); for later rounds,
\begin{equation}
    b_i^{t+1}
    =
    \rho_b\, b_i^t + (1-\rho_b)\,\Delta_{i,\mathrm{safe}}^t,
    \label{eq:b_update}
\end{equation}
where \(\Delta_{i,\mathrm{safe}}^t\) is the aggregation-path message used in
the server history update. The reported experiments use the reconstructed
message \(\tilde{\Delta}_i^t\) for this update:
\begin{equation}
    \Delta_{i,\mathrm{safe}}^t=\tilde{\Delta}_i^t .
    \label{eq:safe_message_bound}
\end{equation}
The uniform bound on this reconstructed message is proved below from the global
cap and the convexity of the residual reconstruction step.

The residual scale follows the implemented EMA recursion
\begin{equation}
    \mu_i^{t+1}
    =
    \rho_\mu\mu_i^t+(1-\rho_\mu)u_i^t,
    \qquad
    0<\rho_\mu<1,
    \label{eq:mu_history_consistency}
\end{equation}
where \(u_i^t\) is the residual-scale input produced by the globally capped
aggregation/statistics path. The implementation initializes new clients with
\(\mu_i^t\le 2B_t\); once \(B_t\le B_{\max}\), the capped-state geometry used
below gives \(0\le u_i^t\le2B_{\max}\). Thus the scale update is a code-level
invariant rather than a distributional assumption.

% In the current implementation used in experiments
% (\texttt{enable\_post\_residual\_b\_cap=False}), no additional post-cap is
% applied to \(\tilde{\Delta}_i^t\) after residual reconstruction.

\subsection{Adaptive weighting}

HRAC assigns weights \(w_i^t \ge 0\) satisfying
\begin{equation}
    \sum_{i=1}^N w_i^t = 1,
    \label{eq:weight_simplex}
\end{equation}
where the weights are determined from the statistics
\(\{\nu_i^t\}_{i=1}^N\).
For the conditional analysis below, \(w_i^t\) is computed from the history
statistics available at the start of round \(t\); the current round updates
only affect the next statistics state.

The final aggregate is
\begin{equation}
    g_t = \sum_{i=1}^N w_i^t \tilde{\Delta}_i^t.
    \label{eq:aggregate}
\end{equation}

\section{Assumptions}

\begin{assumption}[Lower bounded objective]
\label{ass:lower_bounded}
\(f\) is bounded below: \(f(x)\ge f^\star\) for all \(x\in\R^d\).
\end{assumption}

\begin{assumption}[\(L\)-smoothness]
\label{ass:smooth}
\(\norm{\nabla f(x)-\nabla f(y)}\le L\norm{x-y}\) for all \(x,y\in\R^d\).
\end{assumption}

\begin{assumption}[Bias of the honest conditional centre]
\label{ass:bias_variance}
Conditioned on \(\F_t\), the honest conditional centre has bounded bias:
\begin{equation}
    \norm{
        \bar m_{\HH}^t - \nabla f(x_t)
    }
    \le \xi,
    \label{eq:bias}
\end{equation}
\end{assumption}

\begin{assumption}[Bounded honest update magnitude]
\label{ass:honest_update_bound}
There exists \(G_H<\infty\) such that for all honest clients \(i\in\HH\) and
all rounds \(t\),
\begin{equation}
    \norm{\Delta_i^t}\le G_H.
    \label{eq:honest_update_bound}
\end{equation}
\end{assumption}

\begin{assumption}[Honest majority for robust median cap]
\label{ass:honest_majority}
The Byzantine population is strictly less than the honest population:
\begin{equation}
    B < H
    \qquad(\text{equivalently } B < N/2).
    \label{eq:honest_majority}
\end{equation}
\end{assumption}

\begin{assumption}[Honest global clipping admissibility]
\label{ass:global_sadiev}
There exist constants \(\alpha_g\in(1,2]\) and
\(\sigma_{\mathrm{gc}}\ge0\) such that, for every honest client \(i\in\HH\)
and every round \(t\),
\begin{equation}
    \norm{m_i^t}
    \le
    \frac{B_t}{2},
    \label{eq:global_sadiev_admissibility}
\end{equation}
and
\begin{equation}
    \E\!\left[
        \norm{\Delta_i^t-m_i^t}^{\alpha_g}
        \,\middle|\,\F_t
    \right]
    \le
    \sigma_{\mathrm{gc}}^{\alpha_g}.
    \label{eq:global_sadiev_moment}
\end{equation}
With \(B_{\max}:=c_gG_H\), we define
\begin{equation}
    \mathcal C_{\mathrm{gc}}
    :=
    18B_{\max}^{2-\alpha_g}\sigma_{\mathrm{gc}}^{\alpha_g}.
    \label{eq:global_sadiev_budget_constant}
\end{equation}
\end{assumption}

\begin{assumption}[Honest residual clipping admissibility]
\label{ass:residual_sadiev}
For each honest client, define the residual conditional centre
\begin{equation}
    s_i^t:=\E[r_i^t\mid\F_t].
    \label{eq:residual_conditional_centre}
\end{equation}
There exist constants \(\alpha_r\in(1,2]\), \(\sigma_{\mathrm{rc}}\ge0\),
and \(\kappa_\mu<\infty\) such that, for every honest client \(i\in\HH\) and
every round \(t\),
\begin{equation}
    \E\!\left[\norm{r_i^t}\mid\F_t\right]
    \le
    \kappa_\mu\mu_i^t,
    \qquad
    c\ge 2\kappa_\mu,
    \label{eq:residual_scale_tracking}
\end{equation}
and
\begin{equation}
    \E\!\left[
        \norm{r_i^t-s_i^t}^{\alpha_r}
        \,\middle|\,\F_t
    \right]
    \le
    \sigma_{\mathrm{rc}}^{\alpha_r}.
    \label{eq:residual_sadiev_moment}
\end{equation}
With \(B_{\max}:=c_gG_H\) and \(\tau_{\max}:=2cB_{\max}\), define
\begin{equation}
    \mathcal C_{\mathrm{rc}}
    :=
    18\tau_{\max}^{2-\alpha_r}\sigma_{\mathrm{rc}}^{\alpha_r}.
    \label{eq:residual_sadiev_budget_constant}
\end{equation}
\end{assumption}

\begin{remark}[Admissibility and log-checkable scale certificates]
\label{rem:avg_clip_budget}
The half-radius conditions are explicit admissibility assumptions. They are
not automatic consequences of a current-round median; the point of the
history-EMA global scale is to make \(B_t\) predictable, so that the clipped
moment lemma of Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob} applies
conditionally on \(\F_t\). A sufficient global certificate is
\[
    M_m^t:=\max_{i\in\HH}\norm{m_i^t}
    \le \kappa_g\max\{s_t,s_{\min}\},
    \qquad
    c_g\ge 2\kappa_g .
\]
For the residual layer, \eqref{eq:residual_scale_tracking} says that
\(\mu_i^t\) does not underestimate the honest residual scale by more than
\(\kappa_\mu\), with \(c\ge2\kappa_\mu\) providing the corresponding
half-radius margin. Empirically, these requirements can be checked by logging
moving-window ratios between honest update or residual scale proxies and the
EMA scales \(s_t\) and \(\mu_i^t\).
\end{remark}

\begin{assumption}[Uniform Byzantine mass and honest-weight distortion control]
\label{ass:beta_infty}
Define the residual Byzantine mass and the intra-honest weight distortion by
\begin{equation}
    \beta_t:=\sum_{j\in\BB}w_j^t,
    \qquad
    \varepsilon_{w,t}
    :=
    \sum_{i\in\HH}\left|w_i^t-\frac1H\right|.
    \label{eq:beta_epsw_def}
\end{equation}
There exist constants \(\beta_\infty\in[0,1]\) and
\(\varepsilon_{w,\infty}\ge 0\) such that the weights \(w_i^t\) are
\(\F_t\)-measurable and, for every round \(t\),
\begin{align}
    \beta_t &\le \beta_\infty,
    \label{eq:beta_infty}\\
    \varepsilon_{w,t} &\le \varepsilon_{w,\infty},
    \label{eq:epsw_beta_infty}
\end{align}
\end{assumption}

This assumption is not derived automatically from small Byzantine mass. Even
with \(\beta_t=0\), heterogeneous honest clients can receive nonuniform weights
because their \(\nu_i\) histories differ. The theorem therefore keeps honest
weight distortion as its own error-floor term.

\begin{remark}[How to instantiate \(\beta_\infty\)]
\label{rem:beta_infty_instantiation}
Assumption~\ref{ass:beta_infty} is stated abstractly so that the optimisation
theorem stays compatible with the implemented HRAC weighting rule. In an
idealized exponential-weight model,
\[
    w_i^t
    =
    \frac{\exp(-\lambda \nu_i^t)}
         {\sum_{k=1}^N \exp(-\lambda \nu_k^t)},
\]
if the late-stage separation condition
\(
\nu_j^t \ge \nu_i^t+\gamma
\)
holds for all \(j\in\BB\), \(i\in\HH\), then one may take
\begin{equation}
    \beta_\infty
    :=
    \frac{B e^{-\lambda\gamma}}
         {H+B e^{-\lambda\gamma}}.
    \label{eq:beta_infty_explicit}
\end{equation}
Thus the Byzantine weight mass becomes a fully explicit closed constant. The
remaining Byzantine energy factor is supplied pathwise by
Lemma~\ref{lem:mu_tau_bounded}, which gives the processed-message bound
\(\norm{\tilde\Delta_j^t}\le B_{\max}=c_gG_H\).
\end{remark}

\begin{lemma}[Sadiev clipped moment lemma]
\label{lem:clip_sgd_moment}
Let \(X\in\R^d\) be a random vector, let \(x=\E[X\mid\F]\), and define
\(\widehat X=\clip(X,\lambda)\). Following the clipped moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}, if
\[
    \E\!\left[\norm{X-x}^{\alpha}\mid\F\right]\le \sigma^\alpha,
    \qquad
    \norm{x}\le \frac{\lambda}{2},
    \qquad
    \alpha\in(1,2],
\]
then
\begin{align}
    \norm{\E[\widehat X\mid\F]-x}
    &\le
    \frac{2^\alpha\sigma^\alpha}{\lambda^{\alpha-1}},
    \label{eq:clip_sgd_bias}\\
    \E\!\left[\norm{\widehat X-x}^2\mid\F\right]
    &\le
    18\lambda^{2-\alpha}\sigma^\alpha,
    \label{eq:clip_sgd_second}\\
    \E\!\left[
        \norm{\widehat X-\E[\widehat X\mid\F]}^2
        \,\middle|\,\F
    \right]
    &\le
    18\lambda^{2-\alpha}\sigma^\alpha.
    \label{eq:clip_sgd_variance}
\end{align}
\end{lemma}

\begin{proof}
This is Lemma~5.1 of Sadiev et al.~\cite{sadiev2023highprob}, specialized here
to the notation used by HRAC. The same result is proved in their Appendix C as
Lemma C.1.
\end{proof}

\begin{lemma}[Protected-centre clipping distortion]
\label{lem:protected_center_distortion}
Let \(x\in\R^d\), \(X\in\R^d\), and \(\lambda>0\). If \(\norm{x}\le\lambda\),
then
\begin{equation}
    \norm{\clip(X,\lambda)-X}
    =
    (\norm X-\lambda)_+
    \le
    \norm{X-x}.
    \label{eq:protected_center_distortion}
\end{equation}
\end{lemma}

\begin{proof}
If \(\norm X\le\lambda\), both sides give a zero clipping distortion. If
\(\norm X>\lambda\), then
\[
    \norm{\clip(X,\lambda)-X}
    =
    \norm X-\lambda
    \le
    \norm X-\norm x
    \le
    \norm{X-x},
\]
where the last step is the reverse triangle inequality.
\end{proof}

\begin{proposition}[Pointwise clipped-message control from Sadiev admissibility]
\label{prop:clip_sgd_budget_certificate}
Assume that the weights \(w_i^t\) used at round \(t\) are predictable
\((\F_t\)-measurable). Suppose Assumptions~\ref{ass:honest_update_bound},
\ref{ass:honest_majority}, \ref{ass:global_sadiev},
and \ref{ass:residual_sadiev} hold, and HRAC uses the history-state update rules
\eqref{eq:b_update}--\eqref{eq:mu_history_consistency}. Then
Lemma~\ref{lem:clip_sgd_moment}, i.e.,
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}, gives the global
clipped-estimator control
\begin{equation}
    \E\!\left[\norm{\bar\Delta_i^t-m_i^t}^2\mid\F_t\right]
    \le
    \mathcal C_{\mathrm{gc}},
    \label{eq:clip_sgd_gc_control}
\end{equation}
for every honest client \(i\in\HH\). Moreover, the scale-tracking condition
\eqref{eq:residual_scale_tracking} implies the residual Sadiev admissibility
condition
\begin{equation}
    \norm{s_i^t}\le \frac{\tau_i^t}{2},
    \label{eq:residual_sadiev_admissibility}
\end{equation}
and another application of Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}
gives the residual clipped-estimator control
\begin{equation}
    \E\!\left[\norm{\hat r_i^t-s_i^t}^2\mid\F_t\right]
    \le \mathcal C_{\mathrm{rc}}.
    \label{eq:clip_sgd_rc_control}
\end{equation}
Consequently, the two-stage honest clipped message satisfies
\begin{equation}
    \E\!\left[
        \sum_{i\in\HH}w_i^t\norm{\tilde\Delta_i^t-m_i^t}^2
        \,\middle|\,\F_t
    \right]
    \le
    2\mathcal C_{\mathrm{gc}}+2\mathcal C_{\mathrm{rc}}.
    \label{eq:two_stage_honest_control}
\end{equation}
\end{proposition}

\begin{proof}
For the global clipping layer, \(B_t=c_g\max\{s_t,s_{\min}\}\) is
\(\F_t\)-measurable by
\eqref{eq:global_scale_ema}--\eqref{eq:global_cap}. Applying
Lemma~\ref{lem:clip_sgd_moment}, the clipped moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}, with
\[
    (X,x,\lambda,\alpha)
    =
    (\Delta_i^t,m_i^t,B_t,\alpha_g)
\]
and using Assumption~\ref{ass:global_sadiev} gives
\[
    \E\!\left[\norm{\bar\Delta_i^t-m_i^t}^2\mid\F_t\right]
    \le
    18B_t^{2-\alpha_g}\sigma_{\mathrm{gc}}^{\alpha_g}
    \le
    18B_{\max}^{2-\alpha_g}\sigma_{\mathrm{gc}}^{\alpha_g}
    =
    \mathcal C_{\mathrm{gc}}.
\]
where the second inequality uses Corollary~\ref{cor:Bmax_from_honest}.
Because \(w_i^t\) is \(\F_t\)-measurable and \(\sum_{i\in\HH}w_i^t\le1\),
\[
    \E\!\left[
        \sum_{i\in\HH}w_i^t
        \norm{\bar\Delta_i^t-m_i^t}^2
    \right]
    \le \mathcal C_{\mathrm{gc}}.
\]
This is the per-round global clipped-message control used below.

For the residual layer, Jensen's inequality and
\eqref{eq:residual_scale_tracking} give
\[
    \norm{s_i^t}
    =
    \norm{\E[r_i^t\mid\F_t]}
    \le
    \E[\norm{r_i^t}\mid\F_t]
    \le
    \kappa_\mu\mu_i^t
    \le
    \frac{c\mu_i^t}{2}
    =
    \frac{\tau_i^t}{2},
\]
which proves \eqref{eq:residual_sadiev_admissibility}. Since
\(\tau_i^t=c\mu_i^t\) is determined by the history state, it is
\(\F_t\)-measurable. Applying the clipped
moment lemma of Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob},
Lemma~\ref{lem:clip_sgd_moment}, with
\[
    (X,x,\lambda,\alpha)
    =
    (r_i^t,s_i^t,\tau_i^t,\alpha_r)
\]
and using \eqref{eq:residual_sadiev_moment} gives
\[
    \E\!\left[\norm{\hat r_i^t-s_i^t}^2\mid\F_t\right]
    \le
    18(\tau_i^t)^{2-\alpha_r}\sigma_{\mathrm{rc}}^{\alpha_r}
    \le
    18\tau_{\max}^{2-\alpha_r}\sigma_{\mathrm{rc}}^{\alpha_r}
    =
    \mathcal C_{\mathrm{rc}},
\]
where the second inequality uses Lemma~\ref{lem:mu_tau_bounded}.

It remains to connect the two clipped-estimator controls to the actual HRAC
message. Let
\(
    \bar m_i^t:=\E[\bar\Delta_i^t\mid\F_t].
\)
Since \(b_i^t\) is \(\F_t\)-measurable and
\(r_i^t=\bar\Delta_i^t-b_i^t\), we have
\[
    b_i^t+s_i^t
    =
    b_i^t+\E[r_i^t\mid\F_t]
    =
    \E[\bar\Delta_i^t\mid\F_t]
    =
    \bar m_i^t.
\]
Therefore
\[
    \tilde\Delta_i^t-m_i^t
    =
    (\hat r_i^t-s_i^t)+(\bar m_i^t-m_i^t).
\]
Using \(\norm{a+b}^2\le2\norm a^2+2\norm b^2\), Jensen's inequality, and
\eqref{eq:clip_sgd_gc_control},
\[
    \norm{\bar m_i^t-m_i^t}^2
    =
    \norm{\E[\bar\Delta_i^t-m_i^t\mid\F_t]}^2
    \le
    \E[\norm{\bar\Delta_i^t-m_i^t}^2\mid\F_t]
    \le
    \mathcal C_{\mathrm{gc}}.
\]
Combining this bound with \eqref{eq:clip_sgd_rc_control}, multiplying by the
\(\F_t\)-measurable weights, and using \(\sum_{i\in\HH}w_i^t\le1\) proves
\eqref{eq:two_stage_honest_control}. Thus both the global clipping layer and
the residual clipping layer are controlled by the clipped-moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}; the residual admissibility is
supplied by the large-\(c\) scale-tracking condition.
\end{proof}


\section{Basic Properties of HRAC}

\begin{lemma}[Boundedness of globally clipped inputs]
\label{lem:bounded}
For every round \(t\) and every client \(i\),
\begin{equation}
    \norm{\bar{\Delta}_i^t} \le B_t,
    \label{eq:uniform_bound}
\end{equation}
\end{lemma}

\begin{proof}
The bound \(\norm{\bar{\Delta}_i^t} \le B_t\) follows directly from the
definition of global clipping in~\eqref{eq:global_clip}.

\end{proof}

\begin{lemma}[EMA median-scale protection under honest majority]
\label{lem:median_cap_honest}
Under Assumptions~\ref{ass:honest_majority},
\ref{ass:honest_update_bound}, suppose the implementation warm-starts the
global scale from a median observed under the same honest-majority condition,
so \(s_0=M_{\mathrm{init}}\), and choose the fixed floor with
\(0<s_{\min}\le G_H\). Then the round median \(M_t\) in
\eqref{eq:round_median_norm} and the EMA scale \(s_t\) in
\eqref{eq:global_scale_ema} satisfy
\begin{equation}
    M_t\le G_H,
    \qquad
    s_t\le G_H,
    \qquad
    B_t=c_g\max\{s_t,s_{\min}\}\le c_gG_H.
    \label{eq:median_honest_bound}
\end{equation}
\end{lemma}

\begin{proof}
Because \(H>B\), more than half of the \(N\) samples are honest, so every
median order statistic is bracketed by honest samples. In particular, the
median cannot exceed the maximum honest norm. Assumption~\ref{ass:honest_update_bound}
therefore gives \(M_t\le G_H\) for every counted round. The same median
argument applies to the warm-start median \(M_{\mathrm{init}}\), so
\(s_0=M_{\mathrm{init}}\le G_H\). The EMA recursion then gives by induction
\[
    s_{t+1}
    =
    \rho_gs_t+(1-\rho_g)M_t
    \le
    \rho_gG_H+(1-\rho_g)G_H
    =
    G_H.
\]
Finally \(s_{\min}\le G_H\) by the fixed floor choice, so
\(\max\{s_t,s_{\min}\}\le G_H\) and \(B_t\le c_gG_H\).
\end{proof}

\begin{corollary}[Uniform cap bound from honest majority and bounded honest updates]
\label{cor:Bmax_from_honest}
Under Assumptions~\ref{ass:honest_majority},
\ref{ass:honest_update_bound}, with the code-aligned warm-start and
\(0<s_{\min}\le G_H\),
\begin{equation}
    B_t
    \le
    c_g G_H
    =: B_{\max},
    \qquad \forall t.
    \label{eq:Bmax_from_honest}
\end{equation}
\end{corollary}

\begin{proof}
This is the final inequality of Lemma~\ref{lem:median_cap_honest}.
\end{proof}

\begin{lemma}[EMA boundedness for \(\mu_i^t\)]
\label{lem:mu_ema_bounded}
Suppose nonnegative scalars \(s_t\) satisfy \(s_t\le S_{\max}\) for all \(t\),
and \(\mu_{t+1}=\rho_\mu\mu_t+(1-\rho_\mu)s_t\) with \(0<\rho_\mu<1\) and
\(\mu_0\le S_{\max}\). Then \(\mu_t\le S_{\max}\) for all \(t\).
\end{lemma}

\begin{proof}
Induction with convexity:
\[
    \mu_{t+1}
    =
    \rho_\mu\mu_t+(1-\rho_\mu)s_t
    \le
    \rho_\mu S_{\max}+(1-\rho_\mu)S_{\max}
    =
    S_{\max}.
\]
\end{proof}

\begin{lemma}[Bounded history states under bounded global cap]
\label{lem:mu_tau_bounded}
Assume there exists \(B_{\max}<\infty\) such that \(B_t\le B_{\max}\) for all
rounds \(t\). The reported implementation uses
\(\Delta_{i,\mathrm{safe}}^t=\tilde{\Delta}_i^t\) in the history update.
Under the HRAC history-state update rules
\eqref{eq:b_update}--\eqref{eq:mu_history_consistency}, for every client
\(i\) and every \(t\):
\begin{equation}
    \norm{b_i^t}\le B_{\max},
    \qquad
    \norm{r_i^t}\le 2B_{\max},
    \qquad
    \mu_i^t\le 2B_{\max},
    \qquad
    \tau_i^t=c\mu_i^t\le 2cB_{\max}.
    \label{eq:mu_tau_bounds}
\end{equation}
Moreover, the post-processed message is deterministically bounded:
\begin{equation}
    \norm{\tilde\Delta_i^t}\le B_{\max}.
    \label{eq:tilde_delta_det_bound}
\end{equation}
\end{lemma}

\begin{proof}
For \(b_i^t\), the initialization satisfies
\(\norm{b_i^0}\le B_{\max}\) since it is set from a globally clipped
message. Suppose \(\norm{b_i^t}\le B_{\max}\). Because
\(\norm{\bar\Delta_i^t}\le B_t\le B_{\max}\), residual reconstruction has the
form
\[
    \tilde\Delta_i^t
    =
    b_i^t+\hat r_i^t
    =
    (1-\alpha_i^t)b_i^t+\alpha_i^t\bar\Delta_i^t,
    \qquad
    0\le\alpha_i^t\le1 .
\]
Both endpoints of this convex combination lie in the closed
\(B_{\max}\)-ball, so
\(\norm{\tilde\Delta_i^t}\le B_{\max}\). The implemented history update uses
\(\Delta_{i,\mathrm{safe}}^t=\tilde\Delta_i^t\), hence
\[
    b_i^{t+1}
    =
    \rho_b b_i^t+(1-\rho_b)\tilde\Delta_i^t .
\]
Again by convexity, \(\norm{b_i^{t+1}}\le B_{\max}\). Induction gives
\(\norm{b_i^t}\le B_{\max}\) and \(\norm{\tilde\Delta_i^t}\le B_{\max}\) for
all \(t\).

Because \(\norm{\bar\Delta_i^t}\le B_t\le B_{\max}\),
\[
    \norm{r_i^t}
    =
    \norm{\bar\Delta_i^t-b_i^t}
    \le
    \norm{\bar\Delta_i^t}+\norm{b_i^t}
    \le
    2B_{\max}.
\]
The implemented initialization gives
\(\mu_i^t\le2B_t\le2B_{\max}\) when a client first appears. The scale input
\(u_i^t\) in
\eqref{eq:mu_history_consistency} is computed from residuals between globally
capped messages and the bounded history state. Since both references are
bounded by \(B_{\max}\), the residual scale input satisfies
\(0\le u_i^t\le2B_{\max}\). Lemma~\ref{lem:mu_ema_bounded} with
\(S_{\max}=2B_{\max}\) gives \(\mu_i^t\le2B_{\max}\), and hence
\(\tau_i^t=c\mu_i^t\le2cB_{\max}\).
\end{proof}

\begin{lemma}[Aggregate second moment from deterministic cap chain]
\label{lem:g_second_derived}
Under Assumptions~\ref{ass:honest_majority} and
\ref{ass:honest_update_bound}, together with the HRAC history-state update
rules \eqref{eq:b_update}--\eqref{eq:mu_history_consistency}, for every \(t\),
the aggregation output satisfies
\begin{equation}
    \E\!\left[\norm{g_t}^2 \mid \F_t\right]
    \le
    B_{\max}^2,
    \qquad\text{hence}\qquad
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{g_t}^2 \le B_{\max}^2.
    \label{eq:Gg_derived}
\end{equation}
\end{lemma}

\begin{proof}
By Corollary~\ref{cor:Bmax_from_honest}, \(B_t\le B_{\max}\) for all \(t\). 
Lemma~\ref{lem:mu_tau_bounded} then gives
\(\norm{\tilde\Delta_i^t}\le B_{\max}\) deterministically for every
client \(i\).
Write \(g_t=\sum_{i=1}^N w_i^t \tilde{\Delta}_i^t\) with \(\sum_i w_i^t=1\),
\(w_i^t\ge 0\). Jensen's inequality yields:
\[
    \norm{g_t}^2 = \norm{\sum_{i=1}^N w_i^t \tilde{\Delta}_i^t}^2
    \le
    \sum_{i=1}^N w_i^t \norm{\tilde{\Delta}_i^t}^2
    \le
    \sum_{i=1}^N w_i^t B_{\max}^2 = B_{\max}^2.
\]
Taking expectations preserves the bound.
\end{proof}

\begin{lemma}[Residual clipping bias identity]
\label{lem:clip_identity}
For every client \(i\) and round \(t\),
\begin{equation}
    \norm{\hat{r}_i^t - r_i^t}
    =
    (\norm{r_i^t} - \tau_i^t)_+,
    \label{eq:clip_identity}
\end{equation}
where \((a)_+ := \max\{a,0\}\).
As a consequence,
\begin{equation}
    \norm{\tilde{\Delta}_i^t - \bar{\Delta}_i^t}
    =
    (\norm{r_i^t} - \tau_i^t)_+.
    \label{eq:delta_clip_identity}
\end{equation}
\end{lemma}

\begin{proof}
If \(\norm{r_i^t} \le \tau_i^t\), then \(\hat{r}_i^t = r_i^t\), so the
left-hand side is \(0\), equal to
\((\norm{r_i^t}-\tau_i^t)_+\).

If \(\norm{r_i^t} > \tau_i^t\), then
\[
    \hat{r}_i^t = r_i^t \cdot \frac{\tau_i^t}{\norm{r_i^t}},
\]
hence
\[
    \norm{\hat{r}_i^t - r_i^t}
    =
    \left\|
        r_i^t\left(\frac{\tau_i^t}{\norm{r_i^t}} - 1\right)
    \right\|
    =
    \norm{r_i^t} - \tau_i^t.
\]
This proves~\eqref{eq:clip_identity}.
Since
\[
    \tilde{\Delta}_i^t - \bar{\Delta}_i^t
    =
    (b_i^t+\hat{r}_i^t) - (b_i^t+r_i^t)
    =
    \hat{r}_i^t-r_i^t,
\]
we obtain~\eqref{eq:delta_clip_identity}.
\end{proof}

\section{Aggregation Error Analysis}

Define the aggregation error
\begin{equation}
    e_t := g_t - \bar m_{\HH}^t.
    \label{eq:e_t}
\end{equation}

\begin{lemma}[Aggregation error decomposition]
\label{lem:decomp}
The error \(e_t\) admits the decomposition
\begin{equation}
    e_t
    =
    \underbrace{
    \sum_{i \in \HH} w_i^t(\tilde{\Delta}_i^t - m_i^t)
    }_{=:A_t}
    +
    \underbrace{
    \sum_{i \in \HH}
    \left(w_i^t - \frac{1}{H}\right)m_i^t
    }_{=:W_t}
    +
    \underbrace{
    \sum_{j \in \BB} w_j^t \tilde{\Delta}_j^t
    }_{=:M_t}.
    \label{eq:decomposition}
\end{equation}
\end{lemma}

\begin{proof}
From~\eqref{eq:aggregate},
\[
    g_t
    =
    \sum_{i \in \HH} w_i^t \tilde{\Delta}_i^t
    +
    \sum_{j \in \BB} w_j^t \tilde{\Delta}_j^t.
\]
Subtract \(\bar m_{\HH}^t = \frac{1}{H}\sum_{i\in\HH}m_i^t\) and
add/subtract \(\sum_{i\in\HH} w_i^t m_i^t\) to obtain
\[
    g_t - \bar m_{\HH}^t
    =
    \sum_{i \in \HH} w_i^t(\tilde{\Delta}_i^t-m_i^t)
    +
    \sum_{i \in \HH}
    \left(w_i^t-\frac{1}{H}\right)m_i^t
    +
    \sum_{j \in \BB} w_j^t \tilde{\Delta}_j^t.
\]
\end{proof}

\begin{lemma}[Bounding each term of the decomposition]
\label{lem:term_bounds}
\begin{align}
    \norm{A_t}^2
    &\le
    \sum_{i \in \HH} w_i^t \norm{\tilde{\Delta}_i^t-m_i^t}^2,
    \label{eq:A_t_bound}\\
    \norm{W_t}^2
    &\le
    \varepsilon_{w,t}^2 \cdot \max_{i\in\HH}\norm{m_i^t}^2
    \le
    \varepsilon_{w,t}^2 G_H^2,
    \label{eq:W_t_bound}\\
    \norm{M_t}^2
    &\le
    \beta_t \sum_{j\in\BB} w_j^t \norm{\tilde{\Delta}_j^t}^2.
    \label{eq:M_t_bound}
\end{align}
Consequently,
\begin{equation}
    \norm{e_t}^2
    \le
    3\norm{A_t}^2 + 3\norm{W_t}^2 + 3\norm{M_t}^2.
    \label{eq:e_abc}
\end{equation}
\end{lemma}

\begin{proof}
For \(A_t\), Jensen's inequality gives
\[
    \norm{A_t}^2
    =
    \left\|
        \sum_{i \in \HH} w_i^t(\tilde{\Delta}_i^t-m_i^t)
    \right\|^2
    \le
    \sum_{i \in \HH} w_i^t \norm{\tilde{\Delta}_i^t-m_i^t}^2.
\]
More explicitly, since \(\sum_{i\in\HH} w_i^t = 1-\beta_t \le 1\),
weighted Cauchy--Schwarz yields
\[
    \norm{A_t}^2
    \le
    \left(\sum_{i\in\HH} w_i^t\right)
    \left(\sum_{i\in\HH} w_i^t \norm{\tilde{\Delta}_i^t-m_i^t}^2\right)
    \le
    \sum_{i\in\HH} w_i^t \norm{\tilde{\Delta}_i^t-m_i^t}^2.
\]

For \(W_t\),
\[
    \norm{W_t}
    \le
    \sum_{i \in \HH}
    \abs{w_i^t-\frac{1}{H}} \norm{m_i^t}
    \le
    \varepsilon_{w,t} \max_{i\in\HH}\norm{m_i^t}.
\]
Since \(\norm{m_i^t}\le\E[\norm{\Delta_i^t}\mid\F_t]\le G_H\) by
Assumption~\ref{ass:honest_update_bound}, this yields~\eqref{eq:W_t_bound}
after squaring.

For \(M_t\), by Cauchy--Schwarz and \(\sum_{j\in\BB}w_j^t=\beta_t\),
\[
    \norm{M_t}
    \le
    \sum_{j\in\BB} w_j^t \norm{\tilde{\Delta}_j^t}
\]
and therefore
\[
    \norm{M_t}^2
    \le
    \left(\sum_{j\in\BB}w_j^t\right)
    \left(\sum_{j\in\BB}w_j^t\norm{\tilde{\Delta}_j^t}^2\right)
    =
    \beta_t\sum_{j\in\BB}w_j^t\norm{\tilde{\Delta}_j^t}^2.
\]
which is exactly~\eqref{eq:M_t_bound}.

Finally,
\[
    e_t = A_t + W_t + M_t,
\]
so
\[
    \norm{e_t}^2
    \le
    3\norm{A_t}^2 + 3\norm{W_t}^2 + 3\norm{M_t}^2.
\]
\end{proof}

\section{From Honest Conditional Centre to the True Gradient}

\begin{lemma}[Deviation of the honest conditional centre from the true gradient]
\label{lem:honest_gradient}
Under Assumption~\ref{ass:bias_variance},
\begin{equation}
    \norm{\bar m_{\HH}^t - \nabla f(x_t)}^2
    \le
    \xi^2.
    \label{eq:honest_gradient_gap}
\end{equation}
\end{lemma}

\begin{proof}
This is exactly Assumption~\ref{ass:bias_variance}, squared.
\end{proof}

\section{One-Step Descent Analysis}

\begin{lemma}[One-step descent inequality]
\label{lem:descent}
Suppose Assumptions~\ref{ass:lower_bounded} to \ref{ass:honest_majority}
hold.
Then, for any step size \(\eta_t > 0\),
\begin{align}
    \E[f(x_{t+1})]
    \le\;&
    \E[f(x_t)]
    -
    \frac{\eta_t}{2}\E\norm{\nabla f(x_t)}^2
    \nonumber\\
    &+
    \eta_t \E\norm{e_t}^2
    +
    \eta_t\xi^2
    +
    \frac{L\eta_t^2-\eta_t}{2}\E\norm{g_t}^2.
    \label{eq:descent}
\end{align}
In particular, if \(0<\eta_t\le 1/L\), then
\begin{align}
    \E[f(x_{t+1})]
    \le\;&
    \E[f(x_t)]
    -
    \frac{\eta_t}{2}\E\norm{\nabla f(x_t)}^2
    +
    \eta_t \E\norm{e_t}^2
    +
    \eta_t\xi^2.
    \label{eq:descent_stepsize}
\end{align}
\end{lemma}

\begin{proof}
By Assumption~\ref{ass:smooth},
\[
    f(x_{t+1})
    \le
    f(x_t)
    +
    \langle \nabla f(x_t), x_{t+1}-x_t \rangle
    +
    \frac{L}{2}\norm{x_{t+1}-x_t}^2.
\]
Substituting \(x_{t+1}=x_t-\eta_t g_t\) gives
\[
    f(x_{t+1})
    \le
    f(x_t)
    -
    \eta_t \langle \nabla f(x_t), g_t \rangle
    +
    \frac{L\eta_t^2}{2}\norm{g_t}^2.
\]

Using the identity \(-\langle a, b \rangle = -\frac{1}{2}\norm{a}^2 - \frac{1}{2}\norm{b}^2 + \frac{1}{2}\norm{a-b}^2\), we can rewrite the inner product exactly as:
\[
    -\langle \nabla f(x_t), g_t \rangle
    =
    -\frac{1}{2}\norm{\nabla f(x_t)}^2
    - \frac{1}{2}\norm{g_t}^2
    + \frac{1}{2}\norm{g_t - \nabla f(x_t)}^2.
\]
Now, using the decomposition \(g_t = \bar m_{\HH}^t + e_t\), and applying the inequality \(\norm{a+b}^2 \le 2\norm{a}^2 + 2\norm{b}^2\), we bound the distance between the aggregate and the true gradient:
\[
    \norm{g_t - \nabla f(x_t)}^2
    =
    \norm{e_t + \bar m_{\HH}^t - \nabla f(x_t)}^2
    \le
    2\norm{e_t}^2 + 2\norm{\bar m_{\HH}^t - \nabla f(x_t)}^2.
\]
Substituting this back into the smoothness inequality yields:
\[
    f(x_{t+1})
    \le
    f(x_t)
    - \frac{\eta_t}{2}\norm{\nabla f(x_t)}^2
    +
    \frac{L\eta_t^2-\eta_t}{2}\norm{g_t}^2
    + \eta_t\norm{e_t}^2
    + \eta_t\norm{\bar m_{\HH}^t - \nabla f(x_t)}^2.
\]
Taking the conditional expectation \(\E[\cdot \mid \F_t]\) on both sides and applying Lemma~\ref{lem:honest_gradient} for the honest centre deviation yields \eqref{eq:descent}. If \(0<\eta_t\le1/L\), then \((L\eta_t^2-\eta_t)/2\le0\), so dropping the nonpositive term gives \eqref{eq:descent_stepsize}.
\end{proof}

\section{Main Convergence Theorem}

\begin{proposition}[Average aggregation-error budget]
\label{prop:tight_floor_beta2}
Under Assumptions~\ref{ass:honest_update_bound},
\ref{ass:honest_majority}, \ref{ass:global_sadiev},
\ref{ass:residual_sadiev}, and \ref{ass:beta_infty}, together with the HRAC
history-state update rules \eqref{eq:b_update}--\eqref{eq:mu_history_consistency},
the code-aligned global-scale warm-start, and \(0<s_{\min}\le G_H\),
the HRAC
aggregation error satisfies, for every horizon \(T\ge 1\),
\begin{equation}
    \frac1T\sum_{t=0}^{T-1}\E\norm{e_t}^2
    \le
    C_{\mathrm{agg}},
    \label{eq:agg_error_average}
\end{equation}
where
\begin{align}
    C_{\mathrm{clip}}
    &:=
    6\mathcal C_{\mathrm{gc}},
    \label{eq:C_clip}\\
    C_{\mathrm{res}}
    &:=
    6\mathcal C_{\mathrm{rc}},
    \label{eq:C_res}\\
    C_{\mathrm{byz}}
    &:=
    3\varepsilon_{w,\infty}^2G_H^2
    +
    3c_g^2\beta_\infty^2G_H^2,
    \label{eq:C_byz}\\
    C_{\mathrm{agg}}
    &:=
    C_{\mathrm{clip}}+C_{\mathrm{res}}+C_{\mathrm{byz}}.
    \label{eq:C_agg}
\end{align}
\end{proposition}

\begin{proof}
For the honest part of the decomposition, Proposition~\ref{prop:clip_sgd_budget_certificate}
applies the clipped-moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob} once to the predictable global cap and
once to the residual cap certified by the large-\(c\) scale-tracking condition.
Thus, for each round,
\[
    \E\!\left[
        \sum_{i\in\HH} w_i^t\norm{\tilde{\Delta}_i^t-m_i^t}^2
    \right]
    \le
    2\mathcal C_{\mathrm{gc}}+2\mathcal C_{\mathrm{rc}}.
\]
Using Lemma~\ref{lem:term_bounds}, this contributes
\[
    3\E\!\left[
        \sum_{i\in\HH} w_i^t\norm{\tilde{\Delta}_i^t-m_i^t}^2
    \right]
    \le
    C_{\mathrm{clip}}+C_{\mathrm{res}}.
\]
For every \(t\), Assumptions~\ref{ass:honest_update_bound} and
\ref{ass:beta_infty} give
\[
    3\varepsilon_{w,t}^2G_H^2
    \le
    3\varepsilon_{w,\infty}^2G_H^2,
\]
and Corollary~\ref{cor:Bmax_from_honest} together with
Lemma~\ref{lem:mu_tau_bounded} gives
\(\norm{\tilde\Delta_j^t}\le B_{\max}=c_gG_H\), hence
\[
    3\beta_t \sum_{j\in\BB} w_j^t \norm{\tilde{\Delta}_j^t}^2
    \le
    3\beta_t\left(\beta_t c_g^2G_H^2\right)
    \le
    3c_g^2\beta_\infty^2 G_H^2.
\]
Combining these two displays yields \(C_{\mathrm{byz}}\). Finally, apply
Lemma~\ref{lem:term_bounds}, sum over \(t\), and divide by \(T\).
\end{proof}

\begin{theorem}[Convergence to an explicit HRAC constant neighbourhood]
\label{thm:main}
Suppose Assumptions~\ref{ass:lower_bounded} to \ref{ass:beta_infty} hold, and
HRAC uses the history-state update rules
\eqref{eq:b_update}--\eqref{eq:mu_history_consistency}, the code-aligned
global-scale warm-start, and a floor \(0<s_{\min}\le G_H\).
Let the step size be constant and satisfy
\begin{equation}
    0 < \eta \le \frac{1}{L}.
    \label{eq:stepsize}
\end{equation}
Define the explicit HRAC neighbourhood radius
\begin{equation}
    R_{\mathrm{HRAC}}
    :=
    2\xi^2
    +
    2C_{\mathrm{agg}}.
    \label{eq:R_HRAC}
\end{equation}
Then, for every \(T\ge 1\),
\begin{equation}
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \le
    \frac{2(f(x_0)-f^\star)}{\eta T}
    +
    R_{\mathrm{HRAC}}.
    \label{eq:main_full}
\end{equation}
Consequently,
\begin{equation}
    \limsup_{T\to\infty}
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \le
    R_{\mathrm{HRAC}}.
    \label{eq:main_limsup}
\end{equation}
Therefore the averaged squared gradient norm converges to a first-order
stationary neighbourhood with an explicit, time-independent radius.
\end{theorem}

\begin{proof}
Since \(\eta\le1/L\), sum the sharpened one-step descent inequality
\eqref{eq:descent_stepsize} from \(t=0\) to
\(T-1\):
\begin{align*}
    \E[f(x_T)]
    \le\;&
    f(x_0)
    -
    \frac{\eta}{2}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \\
    &+
    \eta\sum_{t=0}^{T-1}\E\norm{e_t}^2
    +
    \eta T\xi^2.
\end{align*}
Since \(f(x_T)\ge f^\star\), rearranging gives
\[
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \le
    \frac{2(f(x_0)-f^\star)}{\eta T}
    +
    \frac{2}{T}\sum_{t=0}^{T-1}\E\norm{e_t}^2
    +
    2\xi^2.
\]
By Proposition~\ref{prop:tight_floor_beta2},
\[
    \frac1T\sum_{t=0}^{T-1}\E\norm{e_t}^2
    \le
    C_{\mathrm{agg}}.
\]
Substituting these average bounds yields
\eqref{eq:main_full}. Taking \(\limsup\) eliminates the vanishing optimization
term \(2(f(x_0)-f^\star)/(\eta T)\) and gives \eqref{eq:main_limsup}.
\end{proof}

\begin{corollary}[Closed-form HRAC constant interval]
\label{cor:refined_beta2_floor}
Under the assumptions of Theorem~\ref{thm:main},
\begin{align}
    R_{\mathrm{HRAC}}
    =\;&
    2\xi^2
    \nonumber\\
    &+
    12\mathcal C_{\mathrm{gc}}
    +
    12\mathcal C_{\mathrm{rc}}
    +
    6\varepsilon_{w,\infty}^2G_H^2
    +
    6c_g^2\beta_\infty^2G_H^2.
    \label{eq:R_HRAC_explicit}
\end{align}
Hence the averaged stationarity measure is asymptotically confined to the
explicit interval
\begin{equation}
    0
    \le
    \limsup_{T\to\infty}
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \le
    R_{\mathrm{HRAC}}.
    \label{eq:HRAC_interval}
\end{equation}
\end{corollary}

\begin{remark}[Same theorem in schematic constant form]
\label{rem:schematic_R_HRAC}
If one prefers the theorem-level presentation
\[
    R_{\mathrm{HRAC}}
    =
    2\xi^2
    +
    C_{\mathrm{clip}}^{\mathrm{eff}}
    +
    C_{\mathrm{res}}^{\mathrm{eff}}
    +
    C_{\mathrm{byz}}^{\mathrm{eff}},
\]
then the effective constants in the present proof are
\[
    C_{\mathrm{clip}}^{\mathrm{eff}}
    := 12\mathcal C_{\mathrm{gc}},\qquad
    C_{\mathrm{res}}^{\mathrm{eff}}
    := 12\mathcal C_{\mathrm{rc}},\qquad
    C_{\mathrm{byz}}^{\mathrm{eff}}
    := 6\varepsilon_{w,\infty}^2G_H^2
       +6c_g^2\beta_\infty^2G_H^2.
\]
The factor \(2\) relative to the raw aggregation-error constants
\eqref{eq:C_clip}--\eqref{eq:C_byz} comes from the descent step
\(
\frac{2}{T}\sum_t \E\norm{e_t}^2
\)
in \eqref{eq:main_full}.
When \(\mathcal C_{\mathrm{gc}}\) and \(\mathcal C_{\mathrm{rc}}\) are
expanded using the clipped-moment constant \(18\) from
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}, the coefficient
in front of the displayed moment terms becomes \(216=12\times18\). Here
\(18\) is inherited from Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob},
while \(12=2\times3\times2\) comes respectively from the final descent
conversion, the three-term aggregation-error decomposition, and the
two-component clipping decomposition in
Proposition~\ref{prop:clip_sgd_budget_certificate}.
\end{remark}

\begin{corollary}[Interpretation as an explicit stationary neighbourhood]
\label{cor:vanishing}
Under the assumptions of Theorem~\ref{thm:main}, HRAC converges sublinearly to
a first-order stationary neighbourhood of radius \(R_{\mathrm{HRAC}}\). In
particular, the long-run averaged squared gradient norm is controlled by a
closed constant composed of:
\begin{enumerate}[label=\arabic*.]
    \item honest conditional-centre bias \(2\xi^2\);
    \item honest global-clipping constant \(12\mathcal C_{\mathrm{gc}}\);
    \item residual Sadiev clipping constant \(12\mathcal C_{\mathrm{rc}}\);
    \item residual Byzantine mass and honest-weight distortion
    \(6\varepsilon_{w,\infty}^2G_H^2
    +6c_g^2\beta_\infty^2G_H^2\).
\end{enumerate}
\end{corollary}

\section{Weight Control from a \texorpdfstring{\(d\)}{d}-Gap Induced
\texorpdfstring{\(\nu\)}{nu}-Gap}
\label{sec:d_gap_induced_nu_weight}

The convergence theorem above only needs a closed upper bound on the Byzantine
weight mass \(\beta_t\). This section closes that bound through HRAC's adaptive
statistics: persistent attack-induced self-inconsistency yields a positive
drift in the residual-motion statistic \(d_i^t\), the EMA recursion accumulates
that drift into a visible late-stage \(\nu\)-gap, and a realised/log-verifiable
\(\nu\)-gap certificate then yields explicit Byzantine weight suppression.


\paragraph{Notation for this section.}
\(\bar m_{\HH}^t=\frac1H\sum_{i\in\HH}m_i^t\) is the honest conditional-centre
field. Let \(a_i^t\) denote the client-level effective update centre tracked by
HRAC's statistics path before residual clipping. For a Byzantine client \(j\),
write
\[
    a_j^t=\bar m_{\HH}^t+\delta_{j,t},
\]
where \(\delta_{j,t}\) is simply the attack-induced displacement relative to
the honest conditional-centre field. The vector \(q_i^t\) denotes the
statistics-path clipped residual of client \(i\), and
\(d_i^t:=\norm{q_i^t-q_i^{t-1}}\) is the per-round residual-motion statistic
whose EMA is \(\nu_i^t\). The implemented excess score derived from \(\nu_i^t\)
is denoted by \(\chi_i^t\), and
\(\beta_t=\sum_{j\in\BB}w_j^t\) is the total Byzantine aggregation weight.

\subsection{Bounded-influence training dynamics}
\label{subsec:bounded_influence_dynamics}

The optimization proof has already reduced HRAC to an honest training field
plus a bounded aggregation perturbation:
\begin{equation}
    g_t
    =
    \bar m_{\HH}^t+e_t,
    \qquad
    x_{t+1}
    =
    x_t-\eta_t(\bar m_{\HH}^t+e_t).
    \label{eq:hrac_training_dynamics}
\end{equation}
Here \(g_t\) is the actual HRAC aggregate used by the server,
\(\bar m_{\HH}^t\) is the average honest conditional update centre,
\(e_t=g_t-\bar m_{\HH}^t\) is the aggregation error bounded in
Proposition~\ref{prop:tight_floor_beta2}, and \(\eta_t\) is the learning rate.
Thus clipping is not used to claim that Byzantine updates disappear. Its role
is to keep their per-round optimization influence finite while honest-majority
training remains the dominant dynamical reference. This is aligned with the
bounded-influence principle used by centered clipping methods for Byzantine
optimization: Karimireddy et al.~\cite[Algorithm~1]{pmlr-v139-karimireddy21a}
clip each worker vector around a reference/history point, schematically
\[
    z_i
    =
    c+\clip(v_i-c,\tau),
\]
where \(v_i\) is worker \(i\)'s submitted vector, \(c\) is the current centre
or history reference, \(\tau\) is the clipping radius, and \(z_i\) is the
message after centered clipping. Thus even if \(v_i\) is Byzantine, the
deviation \(z_i-c\) is bounded by \(\tau\).

The important point is that HRAC's history centre is client-specific. HRAC does
not need a malicious client to be far from a global centre. Instead, it measures
whether the client's residual sequence remains stable relative to its own EMA
history. This is exactly the quantity computed by the code:
\[
    d_i^t
    =
    \norm{q_i^t-q_i^{t-1}},
\]
where \(q_i^t\) is the statistics-path clipped residual for client \(i\).
The following elementary identity explains why a time-varying attack remains
visible even when the malicious client has its own history centre. In an
idealized deterministic notation, let
\[
    b_i^{t+1}
    =
    \rho_b b_i^t+(1-\rho_b)a_i^t,
    \qquad
    e_i^t:=a_i^t-b_i^t.
\]
Then
\begin{equation}
    e_i^{t+1}
    =
    \rho_b e_i^t+(a_i^{t+1}-a_i^t).
    \label{eq:client_history_high_pass}
\end{equation}
Thus the history residual is a stable high-pass trace of the client's own
update increments. For an honest client, \(a_i^{t+1}-a_i^t\) is mainly ordinary
training drift plus stochastic noise. For a Byzantine client
\(a_j^t=\bar m_{\HH}^t+\delta_{j,t}\), the increment contains the additional
attack-field increment
\[
    \delta_{j,t+1}-\delta_{j,t}.
\]
The EMA centre can absorb slowly varying behaviour, but it lags behind a
persistent time-varying attack displacement. After global and residual clipping,
the amplitude of this displacement is bounded, but the remaining temporal
self-inconsistency is precisely what contributes to
\(\norm{q_j^t-q_j^{t-1}}\).

\subsection{A literature-supported route to persistent residual-motion drift}
\label{subsec:diagonal_d_gap}

The source of the \(\nu\)-gap is not a large raw norm. It is a persistent
temporal mismatch between a client's current effective update and the behaviour
that its own EMA history has already absorbed. Two classic attacks make this
mechanism plausible. First, Xie et al.~\cite{xie2020fall} construct an
inner-product manipulation (IPM) attack whose Krum version sets all Byzantine
gradients to
\[
    u_1=\cdots=u_q=-\epsilon\,\bar v,
    \qquad
    \bar v=\frac1{m-q}\sum_{i=1}^{m-q}v_i,
\]
where \(q\) is the number of Byzantine workers, \(m-q\) is the number of
correct workers, \(v_i\) are the correct gradients, \(\bar v\) is their mean,
and \(\epsilon>0\) controls the attack magnitude in the opposite direction.
Identifying \(\bar v\) with the honest centre \(\bar m_{\HH}^t\) gives, in
the notation above,
\[
    a_j^t=-\epsilon\bar m_{\HH}^t,
    \qquad
    \delta_{j,t}^{\mathrm{IPM}}=-(1+\epsilon)\bar m_{\HH}^t.
\]
Thus IPM contributes the attack-field increment
\[
    \delta_{j,t}^{\mathrm{IPM}}-\delta_{j,t-1}^{\mathrm{IPM}}
    =
    -(1+\epsilon)(\bar m_{\HH}^t-\bar m_{\HH}^{t-1}).
\]
When the honest training field changes along the trajectory
\eqref{eq:hrac_training_dynamics}, the malicious history must track a
time-varying negative multiple of that field. In such rounds, the unabsorbed
part can appear in \(\norm{q_j^t-q_j^{t-1}}\).

Second, Baruch et al.~\cite{BarBarGol:19} propose the ALIE coordinate-wise
perturbation
\[
    (p_{\mathrm{mal}})_k=\mu_k-z_{\max}\sigma_k,
\]
where \(\mu_k\) and \(\sigma_k\) are coordinate mean and standard-deviation
estimates, \(z_{\max}\) is the attack's allowed standard-deviation multiplier,
and \((p_{\mathrm{mal}})_k\) is the malicious value submitted on coordinate
\(k\). Writing these round-\(t\) coordinate estimates as
\(\mu_{t,k}^{\mathrm{coord}}\) and \(\sigma_{t,k}^{\mathrm{coord}}\), and
identifying \(\mu_{t,k}^{\mathrm{coord}}\) with the \(k\)-th coordinate of
\(\bar m_{\HH}^t\), the attack displacement is
\[
    (\delta_{j,t}^{\mathrm{ALIE}})_k
    \approx
    -z_{\max}\sigma_{t,k}^{\mathrm{coord}}.
\]
Consequently,
\[
    (\delta_{j,t}^{\mathrm{ALIE}}-\delta_{j,t-1}^{\mathrm{ALIE}})_k
    \approx
    -z_{\max}
    (\sigma_{t,k}^{\mathrm{coord}}-\sigma_{t-1,k}^{\mathrm{coord}}).
\]
Thus changing coordinate scales can induce diagonal attack-field increments
without requiring large update norms. HRAC clipping bounds their amplitude,
while the client-specific EMA history absorbs them only with lag. When this
attack field varies along the training trajectory, the unabsorbed part can
appear in the same full-vector residual-motion statistic that the code uses.

Let \(q_i^t\) denote the statistics-path clipped residual used by HRAC, and
define
\begin{equation}
    d_i^t := \norm{q_i^t-q_i^{t-1}}.
    \label{eq:diagonal_d_def}
\end{equation}
Here \(q_i^t\) is the statistics-path clipped residual for client \(i\), so
\(d_i^t\) measures the full-vector change in that client's residual behaviour
from round \(t-1\) to round \(t\). For group means, write
\begin{equation}
    \bar d_H^t := \frac1H\sum_{i\in\HH}d_i^t,
    \qquad
    \bar d_B^t := \frac1B\sum_{j\in\BB}d_j^t.
    \label{eq:group_d_means}
\end{equation}
Thus \(\bar d_H^t\) and \(\bar d_B^t\) are the honest and Byzantine group
averages of the residual-motion statistic.

\begin{condition}[Persistent group \(d\)-drift]
\label{cond:group_d_gap}
There exist \(T_d<\infty\) and \(\gamma_d>0\) such that, for all \(t\ge T_d\),
\begin{equation}
    \E\!\left[
        \bar d_B^t-\bar d_H^t
        \,\middle|\,\F_{t-1}
    \right]
    \ge
    \gamma_d.
    \label{eq:group_d_gap_condition}
\end{equation}
\end{condition}

Writing \(D_t:=\bar d_B^t-\bar d_H^t\), this is a persistent conditional drift
condition. It does not require every realised round to have a visible positive
gap. Equivalently, one may think of \(D_t=\gamma_d+\varepsilon_t\), where the
noise term \(\varepsilon_t\) can hide the gap early in training, while
\(\E[\varepsilon_t\mid\F_{t-1}]\ge 0\) preserves the positive drift that the
\(\nu\)-EMA accumulates over time.

\begin{proposition}[Persistent \(d\)-drift is accumulated into a \(\nu\)-gap]
\label{prop:d_gap_to_nu_gap}
Suppose \(0<\rho_\nu<1\) and
\begin{equation}
    \nu_i^{t+1}
    =
    \rho_\nu\nu_i^t+(1-\rho_\nu)d_i^t.
    \label{eq:nu_ema_for_d_gap}
\end{equation}
If Condition~\ref{cond:group_d_gap} holds, then with
\[
    z_t:=\bar\nu_B^t-\bar\nu_H^t,
\]
we have, for all \(t\ge T_d\),
\begin{equation}
    \E[z_{t+1}]
    \ge
    \rho_\nu\E[z_t]+(1-\rho_\nu)\gamma_d.
    \label{eq:expected_nu_gap_recursion}
\end{equation}
Consequently, for every \(t\ge T_d\),
\begin{equation}
    \E[z_t]
    \ge
    \rho_\nu^{t-T_d}\E[z_{T_d}]
    +
    \bigl(1-\rho_\nu^{t-T_d}\bigr)\gamma_d,
    \label{eq:expected_nu_gap_unrolled}
\end{equation}
and therefore
\begin{equation}
    \liminf_{t\to\infty}\E[\bar\nu_B^t-\bar\nu_H^t]
    \ge
    \gamma_d.
    \label{eq:expected_nu_gap_liminf}
\end{equation}
\end{proposition}

\begin{proof}
Averaging \eqref{eq:nu_ema_for_d_gap} over Byzantine and honest clients and
subtracting gives
\[
    z_{t+1}
    =
    \rho_\nu z_t
    +
    (1-\rho_\nu)(\bar d_B^t-\bar d_H^t).
\]
Taking expectation and applying
Condition~\ref{cond:group_d_gap} yields
\eqref{eq:expected_nu_gap_recursion}. Unrolling this affine recursion gives
\eqref{eq:expected_nu_gap_unrolled}, and taking \(\liminf\) gives
\eqref{eq:expected_nu_gap_liminf}.
\end{proof}

Thus a literature-supported persistent residual self-inconsistency is
formalized directly as the group \(d\)-drift
Condition~\ref{cond:group_d_gap}, using exactly the full-vector statistic
computed by HRAC. Proposition~\ref{prop:d_gap_to_nu_gap} then shows that
HRAC's EMA statistic accumulates this persistent \(d\)-drift into a late-stage
\(\nu\)-gap. This matches the empirical pattern: the underlying \(d\)-drift can
be present before it is visually obvious, while early stochastic noise and
history warm-up obscure the realised separation. The EMA recursion acts as a
low-pass accumulator, so a small persistent drift becomes clearer in \(\nu\)
after enough rounds.

\begin{remark}[Expectation-level route versus realised weight control]
\label{rem:expected_d_gap_vs_pathwise_nu}
Proposition~\ref{prop:d_gap_to_nu_gap} gives an expectation-level origin for
the group \(\nu\)-gap. The implemented weights, however, are computed from the
realised statistics, not only from their expectations. The weight-control
argument below therefore uses a late-stage realised \(\nu\)-gap certificate.
The expectation-level route can be upgraded to this certificate under an
additional concentration bound or an empirical log certificate showing that the
realised group means and radii stay close to their expectations on the
late-stage window.

When this appendix is read together with the dissertation experiments, the
diagnostic plots provide qualitative realised-trajectory evidence of this kind.
They are not used as a formal numerical certificate in the theorem. In
particular, Figure~\ref{fig:diag-label-flipping},
Figure~\ref{fig:diag-alie}, and Figure~\ref{fig:diag-ipm} report the logged
HRAC residual-motion statistics and weights under label flipping, ALIE, and
IPM, respectively. These plots are used only as realised-trajectory evidence for
Condition~\ref{cond:realized_nu_gap}: on a selected late-stage window, one can
read or compute the group means \(\bar\nu_H^t,\bar\nu_B^t\) and radii
\(r_H^t,r_B^t\), then certify concrete values of
\(\bar\gamma_\nu,r_H,r_B\).
\end{remark}

\begin{definition}[Late-stage \(\nu\) group means and radii]
\label{def:active_group_stats}
For each round \(t\), define
\[
    \bar\nu_H^t := \frac1H\sum_{i\in\HH}\nu_i^t,
    \qquad
    \bar\nu_B^t := \frac1B\sum_{j\in\BB}\nu_j^t.
\]
The within-group \(\nu\)-radii are
\[
    r_H^t := \max_{i\in\HH}\abs{\nu_i^t-\bar\nu_H^t},
    \qquad
    r_B^t := \max_{j\in\BB}\abs{\nu_j^t-\bar\nu_B^t}.
\]
\end{definition}

\begin{condition}[Late-stage realised \(\nu\)-gap certificate]
\label{cond:realized_nu_gap}
There exist \(T_\nu<\infty\), \(\bar\gamma_\nu>0\), and radii
\(r_H,r_B\ge 0\) such that for all \(t\ge T_\nu\),
\begin{equation}
    \bar\nu_B^t-\bar\nu_H^t \ge \bar\gamma_\nu,
    \qquad
    r_H^t\le r_H,
    \qquad
    r_B^t\le r_B.
    \label{eq:active_eventual_mean_nu_gap}
\end{equation}
\end{condition}

\begin{remark}[Logged closing certificate]
\label{rem:logged_closing_certificate}
Condition~\ref{cond:realized_nu_gap} is not an additional algorithmic
mechanism. It is a realised-trajectory closing certificate: on a late-stage
window of the experimental logs, one reads or computes
\(\bar\nu_H^t,\bar\nu_B^t,r_H^t,r_B^t\) and checks that the displayed
inequalities hold with explicit constants. Once certified, these constants
close the remaining gap between the expectation-level \(d\)-drift argument and
the pathwise weight bound used by the implemented HRAC rule.
\end{remark}

\begin{proposition}[Realised mean \(\nu\)-gap gives an implemented score gap]
\label{prop:active_nu_to_score_gap}
Suppose the realised late-stage logs certify
Condition~\ref{cond:realized_nu_gap}. Define
\[
    \bar\nu_t := \frac1N\sum_{k=1}^N\nu_k^t
    =
    \frac{H\bar\nu_H^t+B\bar\nu_B^t}{N},
    \qquad
    \chi_i^t:=(\nu_i^t-\bar\nu_t)_+.
\]
Then for all \(t\ge T_\nu\),
\begin{align}
    \min_{j\in\BB}\chi_j^t
    &\ge
    \chi_B^\nu
    :=
    \left(\frac{H}{N}\bar\gamma_\nu-r_B\right)_+,
    \label{eq:chiB_from_nu}\\
    \max_{i\in\HH}\chi_i^t
    &\le
    \chi_H^\nu
    :=
    \left(r_H-\frac{B}{N}\bar\gamma_\nu\right)_+.
    \label{eq:chiH_from_nu}
\end{align}
\end{proposition}

\begin{proof}
For \(j\in\BB\),
\[
    \nu_j^t-\bar\nu_t
    \ge
    \bar\nu_B^t-r_B
    -
    \frac{H\bar\nu_H^t+B\bar\nu_B^t}{N}
    =
    \frac{H}{N}(\bar\nu_B^t-\bar\nu_H^t)-r_B
    \ge
    \frac{H}{N}\bar\gamma_\nu-r_B.
\]
Taking positive parts gives \eqref{eq:chiB_from_nu}. Similarly, for
\(i\in\HH\),
\[
    \nu_i^t-\bar\nu_t
    \le
    \bar\nu_H^t+r_H
    -
    \frac{H\bar\nu_H^t+B\bar\nu_B^t}{N}
    =
    r_H-\frac{B}{N}(\bar\nu_B^t-\bar\nu_H^t)
    \le
    r_H-\frac{B}{N}\bar\gamma_\nu.
\]
Taking positive parts gives \eqref{eq:chiH_from_nu}.
\end{proof}

\begin{corollary}[Closed Byzantine-mass constant from the realised \(\nu\)-gap]
\label{cor:nu_gap_closed_beta}
This corollary is an idealised late-stage weight-control model. It is not a
line-by-line proof of the implemented weighting rule, which also includes the
\(\nu_i>0.9\) activation threshold, EMA smoothing of the penalty score, and an
iterative maximum-weight cap. Suppose the realised late-stage logs certify
Condition~\ref{cond:realized_nu_gap}, so that the score bounds of
Proposition~\ref{prop:active_nu_to_score_gap} hold. Suppose that, after a
finite time \(T_\star\), the late-stage weights are approximated by the
inactive-cap inverse-score form
\[
    \hat w_i^t=\frac{1}{1+\alpha_w \chi_i^t},
    \qquad
    w_i^t=\frac{\hat w_i^t}{\sum_{k=1}^N\hat w_k^t},
\]
where \(\alpha_w>0\) is the penalty strength, and suppose
\(\chi_B^\nu>\chi_H^\nu\). Define
\[
    T_{\mathrm{close}}:=\max\{T_\star,T_\nu\}.
\]
Then for all \(t\ge T_{\mathrm{close}}\),
\begin{equation}
    \beta_t
    :=
    \sum_{j\in\BB}w_j^t
    \le
    \beta_\infty^\nu
    :=
    \frac{B u_B^\nu}{H u_H^\nu+B u_B^\nu},
    \qquad
    u_B^\nu:=\frac{1}{1+\alpha_w \chi_B^\nu},
    \quad
    u_H^\nu:=\frac{1}{1+\alpha_w \chi_H^\nu}.
    \label{eq:beta_nu_closed}
\end{equation}
If, in addition, for all \(t\ge T_{\mathrm{close}}\),
\[
    \varepsilon_{w,t}\le \varepsilon_{w,\infty}^\nu,
\]
then the Byzantine part of the asymptotic HRAC radius is controlled by
\begin{equation}
    C_{\mathrm{byz}}^\nu
    :=
    3(\varepsilon_{w,\infty}^\nu)^2G_H^2
    +
    3c_g^2(\beta_\infty^\nu)^2G_H^2.
    \label{eq:C_byz_nu_closed}
\end{equation}
\end{corollary}

\begin{proof}
For Byzantine clients, Proposition~\ref{prop:active_nu_to_score_gap} gives
\(\chi_j^t\ge \chi_B^\nu\), hence
\(\hat w_j^t\le u_B^\nu\). For honest clients, it gives
\(\chi_i^t\le \chi_H^\nu\), hence
\(\hat w_i^t\ge u_H^\nu\). Therefore
\[
    \beta_t
    =
    \frac{\sum_{j\in\BB}\hat w_j^t}
         {\sum_{i\in\HH}\hat w_i^t+\sum_{j\in\BB}\hat w_j^t}
    \le
    \frac{B u_B^\nu}{H u_H^\nu+B u_B^\nu}.
\]
The closed Byzantine constant follows by substituting this bound into the
weight-distortion and Byzantine-energy terms of
Lemma~\ref{lem:term_bounds}; the Byzantine energy part uses the pathwise bound
\(\norm{\tilde\Delta_j^t}\le B_{\max}=c_gG_H\) from
Lemma~\ref{lem:mu_tau_bounded}.
\end{proof}

\begin{corollary}[Main HRAC interval closed by Sadiev clipping and the realised \(\nu\)-gap certificate]
\label{cor:sadiev_nu_closed_interval}
Suppose the assumptions of Theorem~\ref{thm:main} hold, except that the
constant \(\beta_\infty\) in Assumption~\ref{ass:beta_infty} is only required
after a finite transient and is supplied by
Corollary~\ref{cor:nu_gap_closed_beta}. Suppose also that the per-round clipping
controls are certified by Proposition~\ref{prop:clip_sgd_budget_certificate}:
the global control by the clipped-moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob} applied to the predictable
history-EMA cap, and the residual control by the same lemma applied
under the large-\(c\) residual scale-tracking condition. Thus
\[
    \mathcal C_{\mathrm{gc}}
    =
    18B_{\max}^{2-\alpha_g}\sigma_{\mathrm{gc}}^{\alpha_g},
    \qquad
    \mathcal C_{\mathrm{rc}}
    =
    18\tau_{\max}^{2-\alpha_r}\sigma_{\mathrm{rc}}^{\alpha_r}.
\]
The coefficients \(216\) in the next display are exactly
\(12\times18\): the factor \(18\) is the second-moment constant in
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob}, and the factor \(12\)
is the proof-level amplification explained in
Remark~\ref{rem:schematic_R_HRAC}.
Then
\begin{align}
    \limsup_{T\to\infty}
    \frac{1}{T}\sum_{t=0}^{T-1}\E\norm{\nabla f(x_t)}^2
    \le\;&
    2\xi^2
    \nonumber\\
    &+
    216B_{\max}^{2-\alpha_g}\sigma_{\mathrm{gc}}^{\alpha_g}
    +
    216\tau_{\max}^{2-\alpha_r}\sigma_{\mathrm{rc}}^{\alpha_r}
    +
    6(\varepsilon_{w,\infty}^\nu)^2G_H^2
    +
    6c_g^2(\beta_\infty^\nu)^2G_H^2.
    \label{eq:sadiev_nu_closed_interval}
\end{align}
\end{corollary}

\begin{proof}
Let \(T_{\mathrm{close}}:=\max\{T_\star,T_\nu\}\). Apply the main descent proof
from \(T_{\mathrm{close}}\) onward. The finite prefix
\(t<T_{\mathrm{close}}\) contributes only \(O(1/T)\) to the averaged
stationarity measure: the prefix length is finite, and the deterministic cap
chain bounds \(g_t\), \(\bar m_{\HH}^t\), and \(e_t\), so its total contribution
is a finite constant divided by \(T\).
Proposition~\ref{prop:clip_sgd_budget_certificate} supplies the two per-round
clipping constants: the global one through the clipped-moment lemma of
Sadiev et al.~\cite[Lemma~5.1]{sadiev2023highprob} applied to
\((X,x,\lambda)=(\Delta_i^t,m_i^t,B_t)\), and the residual one through the
same lemma applied to
\((X,x,\lambda)=(r_i^t,s_i^t,\tau_i^t)\), with admissibility supplied by
\eqref{eq:residual_scale_tracking}. Corollary~\ref{cor:nu_gap_closed_beta}
supplies the closed Byzantine mass constant from the realised late-stage
\(\nu\)-gap certificate. Taking \(\limsup\) gives
\eqref{eq:sadiev_nu_closed_interval}.
\end{proof}


