\doublespacing % Do not change - required

\chapter{Background and Related Work}
\label{ch2}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% IMPORTANT
\begin{spacing}{1}
\minitoc
\end{spacing}
\thesisspacing
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Federated Learning and Data Heterogeneity}

Federated learning (FL) trains a shared model from decentralised data by
iterating between local client optimisation and server-side aggregation.
FedAvg and related model-averaging methods provide the standard starting point
for this setting \cite{DBLP:journals/corr/McMahanMRA16,pmlr-v54-mcmahan17a}.
Clients send training messages to the server rather than raw data, reducing
central data collection and allowing training across distributed devices.
Throughout this thesis, \(\Delta_i^t\) denotes a gradient-like descent
direction submitted by client \(i\) at round \(t\). Thus the server moves in the
negative direction of the aggregate. In the simplest server-side form, mean
aggregation applies
\begin{equation}
    g_t^{\mathrm{mean}} = \frac{1}{N}\sum_{i=1}^{N}\Delta_i^t,
    \qquad
    x_{t+1}=x_t-\eta g_t^{\mathrm{mean}} .
    \label{eq:related-mean}
\end{equation}
This formula is useful but fragile. A single large or consistently biased
\(\Delta_i^t\) can shift the
global direction. Production deployments must handle partial participation,
client availability, stragglers, communication limits, and privacy-preserving
aggregation protocols \cite{bonawitz2019federatedscale,wang2021fieldguide}.
Benchmarking work such as LEAF also emphasises that realistic federated data
are naturally heterogeneous across users and tasks \cite{caldas2018leaf}.

The main complication for robust aggregation is that realistic FL is rarely
IID. Client datasets can be label-skewed and unbalanced, and they are often
shaped by local usage patterns \cite{KairouzMcMahanAvent:21}. As a result,
benign client messages may
legitimately differ in direction and magnitude. Optimisation work such as
FedProx \cite{li2020fedprox}, SCAFFOLD
\cite{DBLP:journals/corr/abs-1910-06378}, and adaptive server methods
\cite{reddi2021adaptive} address benign instability under heterogeneity, but
these methods do not by themselves protect the server from malicious updates.
Empirical studies of federated visual classification further show that the
degree of non-IIDness can materially change FedAvg behaviour
\cite{hsu2019noniidvisual}.
For this paper, heterogeneity is not only an optimisation difficulty, because benign diversity can be mistaken for adversarial
behaviour.
Recent Byzantine-robust FL work therefore treats heterogeneity as part of the
security problem rather than as a secondary nuisance. Resampling and mixing
strategies have been proposed to reduce the gap between benign heterogeneous
clients before applying robust aggregation
\cite{DBLP:journals/corr/abs-2006-09365,pmlr-v206-allouah23a}. HRAC follows
the same broad motivation, but it avoids constructing mixed client batches and
instead keeps a lightweight history for each individual client.

\section{Byzantine-Robust Aggregation}

Byzantine robustness studies aggregation when an unknown subset of workers may
send arbitrary or coordinated messages
\cite{10.1145/357172.357176,guerraoui2024primer}. In distributed and federated learning, the server usually sees only the submitted
updates, not the clients' data or intentions. The aggregation rule is therefore
the main defence against faulty or malicious clients. 

Existing defences can be grouped by their aggregation principle. Selection-based
rules such as Krum choose updates that appear close to other updates
\cite{NIPS2017_f4b9ec30}. In one common description, Krum scores client \(i\)
by the sum of squared distances to its \(N-b-2\) nearest neighbours and selects
the lowest-scoring update:
\begin{equation}
    s_i=\sum_{j\in \mathcal{N}_i}\|\Delta_i^t-\Delta_j^t\|_2^2,
    \qquad
    g_t^{\mathrm{Krum}}=\Delta_{i^\star}^t,\quad
    i^\star=\arg\min_i s_i .
    \label{eq:related-krum}
\end{equation}
Filtering-based methods use a related distance-based idea but remove suspected
outliers before averaging the remaining updates. FABA, for example, repeatedly
removes the update farthest from the current aggregate estimate and then
averages the surviving updates \cite{xia2019faba}.
Coordinate-wise methods such as median and trimmed mean apply robust statistics
independently by coordinate \cite{pmlr-v80-yin18a}. For coordinate \(k\), these
rules can be written as
\begin{equation}
    [g_t^{\mathrm{med}}]_k
    = \operatorname{median}_{i}\,[\Delta_i^t]_k,
    \qquad
    [g_t^{\mathrm{trim}}]_k
    = \frac{1}{N-2b}\sum_{r=b+1}^{N-b} z_{(r),k},
    \label{eq:related-median-trimmed}
\end{equation}
where \(z_{(r),k}\) are the sorted values of
\(\{[\Delta_i^t]_k\}_{i=1}^{N}\).
High-dimensional robust estimation methods, including geometric-median-based
RFA \cite{Pillutla_2022}, estimate a central update with reduced sensitivity to
outliers. Centered clipping instead treats clipping as influence control rather
than attacker removal \cite{pmlr-v139-karimireddy21a}. The Byzantine message may
still enter the average, but only through a bounded residual around a reference
point. Its analysis is most natural when honest updates concentrate around a
common reference, which is closer to an IID or weakly heterogeneous picture
than to strongly label-skewed FL. A typical centred-clipping step clips each
message around a reference \(c_t\):
\begin{equation}
    \tilde{\Delta}_i^t
    = c_t + \min\!\left(1,\frac{\tau}{\|\Delta_i^t-c_t\|_2}\right)
      (\Delta_i^t-c_t),
    \qquad
    g_t=\frac{1}{N}\sum_i \tilde{\Delta}_i^t .
    \label{eq:related-cc}
\end{equation}
Other lines of work directly address robustness under heterogeneous data, such
as RSA, which incorporates a regularisation-based robust aggregation objective
for non-IID distributed datasets \cite{li2019rsa}.
Outside strictly Byzantine settings, adaptive clipping has also been used in
private FL to tune clipping thresholds from update-norm statistics rather than
from expensive manual search \cite{andrew2021adaptiveclipping}. This line of
work supports the use of norm-based, data-dependent caps, although its goal is
privacy accounting rather than adversarial robustness.

Together, these methods show that robustness often comes from reducing the
effect of non-central updates. The limitation is that benign clients may also
be non-central under heterogeneous data. A single centre can therefore be a
poor reference for legitimate label-driven differences. The experiments in
Chapter~\ref{ch:experiments-results} return to this issue for centred
clipping. Selection-based rules also have known vulnerabilities under
coordinated attacks \cite{pmlr-v80-mhamdi18a}.
Variance-based attacks add another difficulty because malicious updates can look
normal under benign update statistics while still damaging training
\cite{BarBarGol:19}. Recent robust-estimation work also notes a
sample-efficiency trade-off, where robust rules may need more participating
clients than averaging to recover the same utility in clean settings
\cite{vlaski2022robustefficientaggregation}. These observations motivate robust methods that limit harmful updates without
automatically penalising every client that differs from the population centre.
\section{Poisoning Attacks and Attack-Dependent Robustness}

Poisoning attacks can target the submitted descent direction directly or
corrupt local data so that protocol-following training produces harmful
messages
\cite{KairouzMcMahanAvent:21}. Model-poisoning attacks are often stronger
because the attacker can craft the submitted message itself. Some attacks are
designed specifically to degrade learning while evading robust aggregation rules
\cite{10.5555/3489212.3489304}. Production-oriented evaluations further show that the damage caused by
poisoning depends on deployment assumptions, threat-model realism, and how the
attack interacts with the aggregation rule \cite{9833647}. This makes it important to report not only final accuracy, but
also the conditions under which a defence is expected to work.

Stealthy attacks are especially relevant to this project because they challenge
simple outlier assumptions. ALIE exploits benign variance so that malicious
updates remain within a statistically plausible range \cite{BarBarGol:19}. If
\(\mu\) and \(\sigma\) are coordinate-wise statistics of the benign updates,
ALIE can submit
\begin{equation}
    \Delta_{\mathrm{ALIE}} = \mu - z\sigma ,
    \label{eq:related-alie}
\end{equation}
for a scalar \(z\) chosen so that the malicious update is shifted but not
obviously extreme in each coordinate. IPM and related sign or direction
manipulations can push the aggregate away from the benign direction without
requiring obviously large norms; a simple IPM form is
\begin{equation}
    \Delta_{\mathrm{IPM}} = -\epsilon\,\bar{\Delta}_{\mathrm{benign}},
    \label{eq:related-ipm}
\end{equation}
where \(\bar{\Delta}_{\mathrm{benign}}\) is the benign mean direction and
\(\epsilon>0\) controls attack strength. Model Shuffle Attack (MSA) further
manipulates parameter structure by shuffling locations and scaling updates,
degrading differential-privacy-based FL while avoiding naive anomaly checks
\cite{YANG2023158}. These attacks suggest that robustness cannot depend only on
detecting extreme update magnitudes.

Data poisoning represents a protocol-compliant and often realistic threat model. In
label flipping, a malicious client changes part of its local labels and then
follows the normal training protocol. The submitted message can therefore look
plausible because it is produced by standard local optimisation rather than by
direct message fabrication. Label-poisoning attacks and defences have been
studied in FL settings
\cite{10.1007/978-3-030-58951-6_24,9897807}.
Formally, the worker still sends a protocol-compliant message, but it is computed
from a poisoned local objective, for example
\begin{equation}
    \tilde{f}_i(x)
    = \frac{1}{|\mathcal{D}_i|}
      \sum_{(u,y)\in \mathcal{D}_i}
      \ell(x;u,\pi(y)),
    \qquad
    \Delta_i^t \approx \nabla \tilde{f}_i(x_t),
    \label{eq:related-label-poisoning}
\end{equation}
where \(\pi\) denotes a label-flipping map. This explains why label poisoning
can be difficult for geometry-only filters: the message is harmful but is still
generated by ordinary local training.

Recent work makes this point sharply for label poisoning: robust aggregators
designed for arbitrary model poisoning can become overly conservative, and may
underperform when the data are sufficiently heterogeneous
\cite{JMLR:v26:24-1307}. Gradient-space methods such as LFighter attempt to
detect poisoned clients by clustering worker gradients \cite{JEBREEL2024111},
but clustering can become unreliable when benign data are strongly non-IID.
HRAC takes two lessons from these results. First, hard filtering should be
avoided when benign heterogeneous updates may be useful. Second, robustness has
to be reported with the attack model and the degree of client heterogeneity,
rather than under a single Byzantine abstraction.

\section{History-Aware and Lightweight Defences}

History is useful in distributed learning because it can stabilise noisy
optimisation. Server-side momentum and adaptive optimisation use temporal state
to improve convergence under stochasticity and partial participation
\cite{reddi2021adaptive}. A representative server momentum recursion is
\begin{equation}
    m_t = \beta m_{t-1} + (1-\beta) g_t,
    \qquad
    x_{t+1}=x_t-\eta m_t .
    \label{eq:related-momentum}
\end{equation}
Robust optimisation can also exploit history: learning-from-history frameworks
construct robust directions from past information, and centered clipping with
momentum demonstrates that historical state can improve Byzantine resilience
\cite{pmlr-v139-karimireddy21a}.
Variance-reduced and momentum-based robust methods similarly argue that
reducing stochastic variation can make malicious messages easier to separate
from noisy benign updates \cite{wu2020byrdsaga,farhadkhani2022resam}.
Historical-gradient studies also examine whether storing past client gradients
helps detect attacks, but they highlight the storage and privacy costs of
keeping high-dimensional historical update records
\cite{choi2024historicalgradients}. This motivates compact temporal summaries
rather than full-gradient histories.

A malicious client may
apply small but persistent perturbations that are inconspicuous in a single
round but accumulate over time. HisMSA explicitly models historical knowledge
to craft poisoning updates that bypass multiple countermeasures
\cite{ZhangHisMSA:2025}. This creates a design tension: a defence should use
history, but it should do so in a way that detects persistent residual change
rather than blindly trusting historical momentum.

Deployability adds a further constraint to history-aware robustness. Practical
FL systems may use privacy-enhancing mechanisms such as secure aggregation. In
this setting, the server recovers only an aggregate sum rather than each
individual update \cite{BonawitzIvanovKreuter:17}. This constrains robust
defences, since many anomaly detectors require detailed per-client visibility.
Other approaches, such as FLTrust, use a server-side reference update to
reweight client messages by alignment \cite{DBLP:journals/corr/abs-2012-13995}.
Such methods improve robustness but introduce additional assumptions about the
server's trusted data or reference direction.

Scalability also matters. Pairwise distances, clustering, and repeated
high-dimensional comparisons can become expensive as the number of clients and
model dimension increase. Lightweight per-client statistics, such as update
norms, residual scales, and residual-change summaries, offer a practical
alternative. They can be maintained with exponential moving averages and used to
detect abrupt or persistent behavioural changes without repeatedly solving a
large robust-estimation problem.

\section{Research Gap}

The related work motivates a narrower problem than simply ``be robust to all
Byzantine clients''. Worst-case robust aggregators can be effective against
arbitrary model poisoning, but under heterogeneous data they may suppress
benign clients that naturally differ from the majority. Mean aggregation has
the opposite weakness. It is less conservative, but it remains exposed to
stronger model-poisoning attacks.

HRAC is positioned between these two behaviours. It keeps an explicit norm cap
for strong malicious updates, but it does not rely only on distance from a
global population centre. This is where it departs from centre-based clipping.
The global cap controls the scale of each submitted message, while the residual
test compares each client mainly with its own historical behaviour. Clients
whose residual dynamics become unstable are penalised softly rather than
removed outright.

