\label{Conclusions}

This project studied Byzantine-robust aggregation under the trade-off between
robustness and heterogeneity tolerance. Mean aggregation is not conservative
enough for strong model-poisoning attacks, while robust aggregation can become
too conservative when benign non-IID clients naturally deviate from the
population.

HRAC addresses this trade-off by limiting the influence of unstable updates. It
combines an EMA-smoothed median norm cap, per-client history-residual clipping,
and delayed soft weighting based on residual-change statistics. The method
compares each client mainly with its own historical behaviour rather than
treating distance from a global centre as sufficient evidence of maliciousness.

On CIFAR-10, HRAC remains competitive in benign training in the reported cached
runs. Its strongest improvements appear under MSA and under non-IID ALIE/IPM. The
diagnostic plots also show strong attacker down-weighting in several settings,
including bit flipping, although the final-accuracy tables show that FABA can
still outperform HRAC in some bit-flipping cases. The ablations support the
component design: the global cap is essential under MSA, while $\nu$-based
temporal weighting is most important under ALIE.

The results also show the method's limits. Label flipping remains more
attack- and partition-dependent, especially when protocol-compliant poisoning is
combined with non-IID data. The reported experiments are controlled cached runs
on CIFAR-10 with \(N=10\), seed 100, \(b=2\) or \(b=3\), and one label class per
client in the non-IID split, so they should be treated as single-seed trends
rather than confidence-interval benchmarks. The next experimental step is not
only to add more runs, but to vary the factors that could change the ranking:
dataset and model family, client count, attacker fraction, label-skew severity,
and attack strength. HRAC also assumes that the server can observe per-client
updates and maintain per-client history statistics, so it is not directly
compatible with standard secure aggregation without additional protocol support.
